ADVANCED CODING

Advanced means:
- powerful but optional to know from memory
- know how to read and use it if provided
- more of a nice-to-have, not necessity


WRITING FUNCTIONS

Formula: x^2 - ln(y)
Write the formula into a function
```{r}
form <- function(x,y){
  x^2-log(y)
}
form(3,5)
form(3:5,8:10)
```

A function running multiple lines

The function in action
```{r}
meanvar <- function(vector){
  m <- mean(vector)
  v <- var(vector)
  return(c(m,v))
}
meanvar(3:5)
```

LOOPS

Check out the iris dataset
```{r}
dim(iris)
head(iris)
```

Objective: vector of cumulative petal lengths
```{r}
c(iris$Petal.Width[1],
  sum(iris$Petal.Length[1:2]),
  sum(iris$Petal.Length[1:3]))
```

for loop: stop at a specified end
```{r}
cpl <- iris$Petal.Length[1]
for (i in 2:150) {
  val <- cpl[i - 1] + iris$Petal.Length[i]
  cpl <- c(cpl, val)
}

cpl
```

while loop: stop at a false condition
```{r}
i <- 1
while (cpl[i] < 10) {
  i <- i + 1
}

i
```


FAMILY OF apply FUNCTIONS

Runs a function repeatedly
```{r}
mean(iris$Sepal.Length)
mean(iris$Petal.Length)
mean(iris$Petal.Width)
```

Objective: average for the numeric columns 

apply
- X
- MARGIN
- FUN
```{r}
apply(X=iris[, -5],MARGIN = 2, FUN=mean)
```

Factors
```{r}
iris$Species
```

Objective: average petal length for each species
```{r}
mean(iris$Petal.Length[iris$Species == "setosa"])
mean(iris$Petal.Length[iris$Species == "versicolor"])
mean(iris$Petal.Length[iris$Species == "virginica"])
```

tapply
- X
- INDEX
- FUN
```{r}
tapply(iris$Petal.Length, iris$Species, FUN = mean)
```
