IFS

Objective: output a message whether x is 5 or not


Use if...else in writing a function that finds the fourth root
```{r}
frrt <- function(b) {
  if (b >= 0) {
    return(b^(1 / 4))
  } else {
    return("Please enter a non-negative number, thanks ;)")
  }
}
```

```{r}
# 5 if versicolor, 1 otherwise
x <- 10/3
if (x ==5) {
  print("x is 5")
} else{
  print("x is not 5")
}
```

Objective: create a new vector from Species
```{r}
# 5 if versicolor, 1 otherwise
ifelse(iris$Species =="versicolor",5,1)
```

MISC MANIP 1 - Counting
```{r}
F+T
T+T
sum(1:8 >=5)
```

How many irises have a petal length shorter than 2?
```{r}
sum(iris$Petal.Length < 2)
```

MISC MANIP 2 - Matching
```{r}
(c("A","B","C","D") =="B") |(c("A","B","C","D") =="D")
```

How many irises have a petal length of 1 or 2 or ... or 6?
```{r}
sum(iris$Petal.Length %in% 1:6)
```

MISC MANIP 3 - paste

Setup
```{r}
new.iris <-iris
new.iris$oddeven <- c("O","E")
```

Merging character objects
```{r}
paste("a","b","c")
paste("a","b","c", sep="+")
paste0("a","b","c")

paste(1:3, collapse = "+")

paste(new.iris$Species, new.iris$oddeven, sep = "-")
```

PIPES

A nesting nightmare
```{r}
sd(apply(select(iris, -Species), MARGIN = 2, FUN = mean))
```

Loading package
```{r}
library(dplyr)
```

select: grab columns
- first argument = data frame
- following arguments = column names; start names with - to drop columns instead


A better way
```{r}
 iris %>% #ctrl+shif+m
  select(-Species) %>% 
  apply(MARGIN = 2, FUN = mean) %>% 
  sd()
```


