VECTORS
```{r}
# A numeric vector
c(1,4,6,2,5)

# A character vector
c("nice","2", "meet")
```

Working with a numeric vector
```{r}
vect <- c(1,4,6,3,2,6,4)
vect
sqrt(vect)
mean(vect)
```

Grabbing from a vector
```{r}
# grab the 5th value
vect[5]

# grab the 2nd and 4th value
vect[c(2,4)]

# grab all except for the 2nd and 4th value
vect[-c(2,4)]

# grab the last and 3rd value twice
vect[c(7,3,3)]

# Quiz
vect[vect]
```

Create sequence of consecutive integers
```{r}
1:10
7:-7
```

DATA FRAMES
```{r}
dataset <- data.frame(Nums=6:10, Group=c("A","B","B","A","B"))
```

Column referencing
```{r}
dataset$Nums
dataset$word <- c("the","quick","brown","fox","jump")
```

Grabbing from a data frame
```{r}
dataset[1,4]
dataset[2,]
dataset[,"Group"]
dataset[2:3,c(1,3)]
```

LOGICAL VECTORS
```{r}
6>10
vect < 4
```

Logic symbols:
  greater than >
  less than <
  at least >=
  at most =<
  equals ==
  not equals !=
  union |
  intersection &
  complement !
  
"=" versus "=="
```{r}
# Assigning name
hello=5

# Checking logic
hello==7
```

Grabbing using logical vectors
Helpful Rule: logical vector should have a matching length
```{r}
# Grab the 4th value
vect[4]
vect[c(F,F,F,T,F,F,F)]

# Grab the values that are not 4
vect[vect!=4]

# Quiz
dataset[dataset$Group=="A",1]
```

Recycling shorter vectors
```{r}
1:4+4  #Ssame as 1:4+C(4,4,4,4)
1:4+4:1
```
