REVISITING "STUDENT DATA: PART 1"

Read the data
```{r}
student <- read.csv(file = "filepath/student_grades.csv", stringsAsFactors = T)
```

Quick peek
```{r}
summary(student)
```

Loading package
```{r}
library(dplyr)  # or library(tidyverse)
```

select: grab columns
- specify column names; start names with - to drop columns instead

filter: grab rows
- specify logical statements

mutate: create a new column
- specify new column's name and formula

arrange: sort by a column
- specify column name for ascending; wrap column name in desc() for descending

Without pipes
```{r}
select(arrange(mutate(filter(student, G3 < 5), grade.diff = G3 - G2), grade.diff), grade.diff, address, reason)
```

With pipes
```{r}
student %>% 
  filter(G3 < 5) %>% 
  mutate(grade.diff = G3 - G2) %>% 
  arrange(grade.diff) %>% 
  select(grade.diff, address, reason)
```

Minor tweaks
```{r}
# Keep observations with G3 between 0 and 20
student2 <- student[(student$G3 >= 0) & (student$G3 <= 20), ]
summary(student2$G3)





# Define new variable: average of G1, G2, and G3
student2$newgrade <- (student2$G1 + student2$G2 + student2$G3) / 3


```

Objective: explore variables newgrade, absences, and Mjob
```{r}
# Univariate
summary(student2$newgrade)
summary(student2$absences)
summary(student2$Mjob)



library(ggplot2)
ggplot(data = student2, mapping = aes(x = newgrade)) +
  geom_histogram()

ggplot(data = student2, mapping = aes(x = absences)) +
  geom_histogram()

# Bivariate
cor(student2$newgrade, student2$absences)

ggplot(data = student2, mapping = aes(x = absences, y = newgrade)) +
  geom_point()

library(dplyr)
student2 %>%
  group_by(Mjob) %>%
  summarise(Means   = mean(newgrade),
            Medians = median(newgrade),
            Freq    = n()) 

ggplot(data = student2, mapping = aes(x = Mjob, y = newgrade)) +
  geom_boxplot()
```

